---
title: "Women's world cup 2023"
author:
  - Ian Hamilton
  - Stefan Stein
  - David Selby
date: July 2023
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
library(parallel)
```



```{r}
First_round_odds <- read_excel("data/First round odds.xlsx") %>%
  rename(
    "odds_1" = "Home win",
    "odds_draw" = "Draw",
    "odds_2" = "Away win",
    "TeamName1" = "Home",
    "TeamName2" = "Away"
  ) %>%
  mutate(TeamName2 = ifelse(TeamName2 == "Republic of Ireland", "Ireland", TeamName2))%>%
  mutate(
    p_ij = 1 / odds_1,
    p_draw = 1 / odds_draw,
    p_ji = 1 / odds_2,
    total_prob = p_ij + p_draw + p_ji
  ) %>%
  mutate(p_ij = p_ij / total_prob,
         p_draw = p_draw / total_prob,
         p_ji = p_ji / total_prob) %>%
  select(-total_prob) %>%
  mutate(s_ij = p_ij / p_ji)
```

```{r}
p_ij <- First_round_odds[1,"p_ij"]%>%as.numeric()
p_draw <- First_round_odds[1,"p_draw"]%>%as.numeric()
s_ij <- First_round_odds[1,"s_ij"]%>%as.numeric()
```


```{r}
target_function_beat <- function(p_ij,s_ij,nu){
  s_ij / (s_ij + 1 + nu*sqrt(s_ij)) - p_ij
}
target_function_draw <- function(p_draw,s_ij,nu){
  nu*sqrt(s_ij) / (s_ij + 1 + nu*sqrt(s_ij)) - p_draw
}
root_beat <- uniroot(target_function_beat, p_ij=p_ij,s_ij=s_ij,interval = c(-1,1))
root_draw <- uniroot(target_function_draw, p_draw=p_draw,s_ij=s_ij,interval = c(-1,1))
nu_fn_beat <- function(p_ij, s_ij){
  uniroot(target_function_beat, p_ij=p_ij,s_ij=s_ij,interval = c(-1,1))[['root']]
}
nu_fn_draw <- function(p_draw, s_ij){
  uniroot(target_function_draw, p_draw=p_draw,s_ij=s_ij,interval = c(-1,1))[['root']]
}
```



```{r}
df <- First_round_odds %>%
  rowwise() %>%
  mutate(nu_beat = nu_fn_beat(p_ij,s_ij),
         nu_draw = nu_fn_draw(p_draw,s_ij)) %>%
  ungroup()
df
```

```{r}
summary(df$nu_beat)
```

```{r}
# based on 10 matches
odds_threshold <- 1.1
df%>%filter(odds_1 >= odds_threshold,
            odds_2 >= odds_threshold)%>%
  pull(nu_beat)%>%
  summary()
```

```{r}
nu <- df%>%filter(odds_1 >= odds_threshold,
            odds_2 >= odds_threshold)%>%
  pull(nu_beat)%>%
  summary()

nu <- nu["Mean"]
nu
```



```{r}
# based on 7 matches
odds_threshold <- 1.3
df%>%filter(odds_1 >= odds_threshold,
            odds_2 >= odds_threshold)%>%
  pull(nu_beat)%>%
  summary()
```


# Fifa rankings

## Getting sigma

```{r}
Fifa_rating <- read_csv("data/Fifa rating.csv")
```



```{r}
df <- df%>%
  left_join(Fifa_rating%>%rename("F_i" = "Fifa rating"),by=c("TeamName1"="Team"))%>%
  left_join(Fifa_rating%>%rename("F_j" = "Fifa rating"),by=c("TeamName2"="Team"))
```

```{r}
df <- df%>%
  mutate(sigma = (F_i - F_j) / log10((p_ji + 0.5*p_draw)/(p_ij + 0.5*p_draw)))
```

```{r}
summary(df$sigma)
```

```{r}
summary(df$sigma[1:15])
```

```{r}
plot(df$sigma[1:15])
```

Here I am doing a regression of

$$
F_i - F_j = \log \left(\frac{p_{ji}+0.5 p_{i \sim j}}{p_{ij}+0.5p_{i \sim j}} \right)\sigma'
$$
to get an estimate of $\sigma'$.

```{r}
M <- df%>%
  transmute(X = log10((p_ji + 0.5*p_draw)/(p_ij + 0.5*p_draw)),
            Y = (F_i - F_j))

M <- M[1:15,]

lmfit <- lm(Y ~ -1 + X, data = M)
sigma_dash <- coef(lmfit)
summary(lmfit)
```

```{r}
plot(lmfit$residuals)
```

## Getting alpha

```{r}
Group_Winner_odds <- read_excel("data/Group Winner odds.xlsx")
```

```{r}

get_group_matches <- function(group){
  Group_Winner_odds%>%
  filter(Group == group)%>%
  pull(Team)%>%
  combn(m=2)%>%
  t()%>%
  as.data.frame()%>%
  mutate(group = group)
}
groups <- LETTERS[1:8]

group_matches <- lapply(groups, get_group_matches)%>%
  do.call(rbind, .)%>%
  rename("TeamName1" = "V1", "TeamName2" = "V2") %>%
  rowwise() %>%
  mutate(match_string = sort(c(TeamName1,TeamName2))%>%paste(collapse = "-"))%>%
  ungroup()
```

```{r}

find_sij <- function(alpha_ij, s_ij){
  (s_ij + 0.5 * nu * sqrt(s_ij)) / (s_ij + 1 + nu*sqrt(s_ij)) - alpha_ij
}
sij_fn <- function(alpha_ij){
  uniroot(find_sij, alpha_ij=alpha_ij,interval = c(0,10000))[['root']]
}

  
estimated_df <- group_matches%>%
  select(-c(TeamName1,TeamName2))%>%
  left_join(
    df%>%
      rowwise()%>%
      mutate(match_string = sort(c(TeamName1,TeamName2))%>%paste(collapse = "-"))%>%
      ungroup(),
    by = "match_string"
  )%>%
  rowwise()%>%
  mutate(TeamName1 = ifelse(is.na(TeamName1), str_split(match_string, pattern="-", simplify = T)[1], TeamName1),
         TeamName2 = ifelse(is.na(TeamName2), str_split(match_string, pattern="-", simplify = T)[2], TeamName2))%>%
  ungroup()%>%
  # drop Fifa rankings here to join them back for all matches in next step
  select(-c(F_i,F_j, sigma))%>%
  select(-starts_with("nu"))%>%
  left_join(Fifa_rating%>%rename("F_i" = "Fifa rating"),by=c("TeamName1"="Team"))%>%
  left_join(Fifa_rating%>%rename("F_j" = "Fifa rating"),by=c("TeamName2"="Team"))%>%
  mutate(alpha_ij = 1 / (1 + 10^((F_i - F_j)/ sigma_dash))) %>%
  rowwise()%>%
  mutate(s_ij = ifelse(is.na(s_ij),sij_fn(alpha_ij),s_ij))%>%
  ungroup()%>%
  # now add the missing probs
  mutate(p_ij = ifelse(is.na(p_ij),s_ij/(s_ij + 1 + nu*sqrt(s_ij)),p_ij),
         p_draw = ifelse(is.na(p_draw),nu*sqrt(s_ij)/(s_ij + 1 + nu*sqrt(s_ij)),p_draw),
         p_ji = ifelse(is.na(p_ji),1-p_ij-p_draw,p_ji),
         total_prob = p_ij + p_draw + p_ji,
         logit_p_ij = log(p_ij / p_ji))

```

```{r}
estimated_df
```

## Investigate the extreme s_ij values

```{r}
fifa_diffs <- seq(-750, 750, by=1)
root_fn <- function(alpha, s){
  (1-alpha)*s+nu*(0.5-alpha)*sqrt(s)-alpha
}
find_y <- function(d){
  alpha <- 1 / (1 + 10^((d)/sigma_dash))
  uniroot(root_fn, alpha=alpha, interval = c(0,10000))[["root"]]
}
ys <- sapply(fifa_diffs, find_y)
plot(fifa_diffs, ys, log = "y", type = "l")
```

```{r}
estimated_df%>%
  mutate(fifa_diff = F_i - F_j)%>%
  ggplot(aes(x=fifa_diff,y=s_ij)) +
  geom_point()
```

# Intra-group strength

We then estimate the intra-group strength parameters by linear regression, starting with constructing an appropriate adjacency matrix,


```{r}
# I'm using Group_Winner_odds$Team here, because that's a list of all the teams

makeX_lm <- function(df){
  X_lm <- matrix(0,
                 nrow(df),
                 length(levels(factor(as.character(Group_Winner_odds$Team)))))
  colnames(X_lm) <- levels(factor(as.character(Group_Winner_odds$Team)))
  for (team in colnames(X_lm)) {
    X_lm[df$TeamName1 == team, team] <- 1
    X_lm[df$TeamName2 == team, team] <- -1
  }
  return(X_lm)
}

estimated_df$X <- makeX_lm(estimated_df)
```

then fitting the linear model.

```{r}
lmfit_group <- lm(logit_p_ij ~ -1 + X, data = estimated_df)

r <- coef(lmfit_group)
r[is.na(r)] <- 0
s <- exp(r)
sort(r)
```

Under Bradley--Terry, the probability of $i$ winning the final, given $i$ has made it to the final, is
\[
q_{io} = \pi_i / (\pi_i + \pi_o),
\]
where $\pi_i$ is the strength of team $i$ and $\pi_o$ is the strength of the final opponent, since there are so many permutations by which a final opponent may be determined for all teams.
The opponent strength $\pi_o$ is assumed to be the same for all teams, so
\begin{equation}
\text{logit}(q_{io}) = \lambda_i - \lambda_o
\end{equation}
where $\lambda_i = \log \pi_i$.
We have the intra-group strengths $s_i$, but we do not know the relative strengths of the groups.
Suppose group $G$ has strength $\gamma_G$ such that $s_i = \gamma_{G(i)} \pi_i$ where $G(i)$ is $i$'s group.
Then
\begin{equation}
\lambda_i = \log \pi_i = \log s_i - \log \gamma_{G(i)} = r_i - \lambda_{G(i)}
\end{equation}
where $\lambda_{G(i)} = \log \gamma_{G(i)}$.
So combining (1) and (2) we have
\[
r_i - \text{logit}(q_{io}) = \lambda_{G(i)} + \lambda_o.
\]
So we estimate $\lambda_o$ and the $\lambda_G$ by regressing against $r_i - \text{logit}(q_{io})$.

```{r}

winner <- read_csv("data/Winner odds.csv")%>%
  # filter(Odds < 100)%>%
  mutate(pi_win = 1 / Odds,
         total_prob = sum(pi_win))%>%
  mutate(pi_win = pi_win / total_prob)%>%
  select(Team, pi_win)%>%
  left_join(
    read_csv("data/Reach final odds.csv")%>%
      mutate(p_i = 1 / Odds,
             total_prob = sum(p_i))%>%
      mutate(p_i = 2*p_i / total_prob)%>%
      select(Team, p_i)%>%
      rename("pi_final" = "p_i"),
    by = "Team"
  )%>%
  # calculate the conditional probability of winning given you make the final
  mutate(q_io = pi_win / pi_final)%>%
  left_join(
    as.data.frame(r)%>%
      rownames_to_column("Team")%>%
      mutate(Team = str_remove(Team, "^X")),
    by = "Team"
  )%>%
  left_join(
    Group_Winner_odds%>%select(-Odds), by="Team"
  )

winner
```

```{r}
winner%>%
  mutate(label = ifelse(Team %in% c("USA", "Brazil", "Germany", "England", "Australia", "Spain"), Team, "")) %>%
  ggplot(aes(x=pi_win, y=q_io, label = label)) +
  geom_point() +
  geom_vline(xintercept=0.03) + 
    labs(x="P[winning the tournament]",
         y="P[winning the tournament, given reaching final]") +
  geom_text(check_overlap = TRUE, vjust=-0.5)
```




```{r}

y <- winner$r - qlogis(winner$q_io)

X <- matrix(0,
            nrow(winner),
            length(unique(winner$Group)))
colnames(X) <- sort(unique(winner$Group))
rownames(X) <- winner$Team
for (group in colnames(X)) {
  X[winner$Group == group, group] <- 1
}
X

lmfit_win <- lm(y ~ X)
```


Where probability of winning is small, we make an adjustment according to a threshold, since the $q_io$ for low probability teams may be highly unreliable.

```{r}
cut <- subset(winner, pi_win > 0.02)

y2 <- cut$r - qlogis(cut$q_io)
X2 <- matrix(0,
            nrow(cut),
            # for columns I still need to take winner df, to ensure all groups are there
            length(unique(winner$Group)))
colnames(X2) <- sort(unique(winner$Group))
rownames(X2) <- cut$Team
for (group in colnames(X2)) {
  X2[cut$Group == group, group] <- 1
}
X2

lmfit_win <- lm(y2 ~ X2) # NOTE: this replaces lmfit_win above
```


```{r}
lambda_G <- coef(lmfit_win)
lambda_G[is.na(lambda_G)] <- 0
names(lambda_G) <- c("O", "A", "B", "C", "D", "E", "F", "G", "H")
lambda_G
```

Get the team coefficients $\lambda_i = r_i - \lambda_{G(i)}$

```{r}
lambda <- setNames(winner$r - lambda_G[winner$Group], winner$Team)
lambda
```


Then calculate the probabilities from the $\lambda$s

```{r}
probs <- plogis(outer(lambda, lambda, '-'))
probs
```

```{r}
write.csv(probs, "probabilities.csv")
```




# Simulation

```{r}
# simulate the group stage given the group name
simulate_group <- function(group_letter) {
  probs <- estimated_df%>%
    filter(group == group_letter)%>%
    rowwise()%>%
    mutate(result = sample(c(3,-3,1),size=1,prob = c(p_ij,p_ji,p_draw)))
  
  group_winners <- probs%>%select(TeamName1, result)%>%rename("team" = "TeamName1")%>%
    rbind(probs%>%select(TeamName2, result)%>%rename("team" = "TeamName2")%>%mutate(result = ifelse(abs(result)>1, -1*result, result)))%>%
    ungroup()%>%
    filter(result > 0)%>%
    group_by(team)%>%
    summarise(points = sum(result))%>%
    arrange(desc(points))%>%
    head(2)%>%
    pull(team)
  return(group_winners)
}


simulate_match <- function(team1,team2){
  p_ij <- probs[team1,team2]
  sample(c(team1,team2),size = 1, prob = c(p_ij, 1 - p_ij))
}

simulate_tournament <- function(x){
  
  # group stage
  winners_A <- simulate_group("A")
  winners_B <- simulate_group("B")
  winners_C <- simulate_group("C")
  winners_D <- simulate_group("D")
  winners_E <- simulate_group("E")
  winners_F <- simulate_group("F")
  winners_G <- simulate_group("G")
  winners_H <- simulate_group("H")
  
  # round of 16
  r16_1 <- simulate_match(winners_A[1],winners_B[2])
  r16_2 <- simulate_match(winners_C[1],winners_D[2])
  r16_3 <- simulate_match(winners_E[1],winners_F[2])
  r16_4 <- simulate_match(winners_G[1],winners_H[2])
  r16_5 <- simulate_match(winners_B[1],winners_A[2])
  r16_6 <- simulate_match(winners_D[1],winners_C[2])
  r16_7 <- simulate_match(winners_F[1],winners_E[2])
  r16_8 <- simulate_match(winners_H[1],winners_G[2])
  
  # round of 8
  r8_1 <- simulate_match(r16_1,r16_2)
  r8_2 <- simulate_match(r16_3,r16_4)
  r8_3 <- simulate_match(r16_5,r16_6)
  r8_4 <- simulate_match(r16_7,r16_8)
  
  # semi-finals
  r4_1 <- simulate_match(r8_1,r8_2)
  r4_2 <- simulate_match(r8_3,r8_4)
  
  final <- simulate_match(r4_1,r4_2)
  
  return(final)
}

```



```{r}
set.seed(1234)
M <- 10
sim_results <- mclapply(1:M,FUN=simulate_tournament,mc.cores = 4)
simplify2array(sim_results)
```



```{r}
library(doParallel)
registerDoParallel(4)
```


```{r}
set.seed(1234)
loop_group_files <- c("group_winners1.csv", "group_winners2.csv", "group_winners3.csv", "group_winners4.csv")
loop_winners_files <- c("winners1.csv", "winners2.csv", "winners3.csv", "winners4.csv")
M <- 2500
results_list <- foreach(i = 1:4) %dopar% {
  group_file <- loop_group_files[i]
  winners_file <- loop_winners_files[i]
  for (k in 1:M) {
    # group stage
    winners_A <- simulate_group("A")
    winners_B <- simulate_group("B")
    winners_C <- simulate_group("C")
    winners_D <- simulate_group("D")
    winners_E <- simulate_group("E")
    winners_F <- simulate_group("F")
    winners_G <- simulate_group("G")
    winners_H <- simulate_group("H")
    
    # round of 16
    r16_1 <- simulate_match(winners_A[1],winners_B[2])
    r16_2 <- simulate_match(winners_C[1],winners_D[2])
    r16_3 <- simulate_match(winners_E[1],winners_F[2])
    r16_4 <- simulate_match(winners_G[1],winners_H[2])
    r16_5 <- simulate_match(winners_B[1],winners_A[2])
    r16_6 <- simulate_match(winners_D[1],winners_C[2])
    r16_7 <- simulate_match(winners_F[1],winners_E[2])
    r16_8 <- simulate_match(winners_H[1],winners_G[2])
    
    # round of 8
    r8_1 <- simulate_match(r16_1,r16_2)
    r8_2 <- simulate_match(r16_3,r16_4)
    r8_3 <- simulate_match(r16_5,r16_6)
    r8_4 <- simulate_match(r16_7,r16_8)
    
    # semi-finals
    r4_1 <- simulate_match(r8_1,r8_2)
    r4_2 <- simulate_match(r8_3,r8_4)
    
    final <- simulate_match(r4_1,r4_2)
    
    readr::write_csv(as.tibble(t(c(k,"A",winners_A)))%>%setNames(c("iteration","Group","First", "Second")), 
                     file = paste0("results_10000/",group_file), append = TRUE)
    readr::write_csv(as.tibble(t(c(k,"B",winners_B)))%>%setNames(c("iteration","Group","First", "Second")),
                     file = paste0("results_10000/",group_file), append = TRUE)
    readr::write_csv(as.tibble(t(c(k,"C",winners_C)))%>%setNames(c("iteration","Group","First", "Second")),
                     file = paste0("results_10000/",group_file), append = TRUE)
    readr::write_csv(as.tibble(t(c(k,"D",winners_D)))%>%setNames(c("iteration","Group","First", "Second")),
                     file = paste0("results_10000/",group_file), append = TRUE)
    readr::write_csv(as.tibble(t(c(k,"E",winners_E)))%>%setNames(c("iteration","Group","First", "Second")),
                     file = paste0("results_10000/",group_file), append = TRUE)
    readr::write_csv(as.tibble(t(c(k,"F",winners_F)))%>%setNames(c("iteration","Group","First", "Second")),
                     file = paste0("results_10000/",group_file), append = TRUE)
    readr::write_csv(as.tibble(t(c(k,"G",winners_G)))%>%setNames(c("iteration","Group","First", "Second")),
                     file = paste0("results_10000/",group_file), append = TRUE)
    readr::write_csv(as.tibble(t(c(k,"H",winners_H)))%>%setNames(c("iteration","Group","First", "Second")),
                     file = paste0("results_10000/",group_file), append = TRUE)
    
    readr::write_csv(as.tibble(final)%>%setNames(c("Winner")),
                     file = paste0("results_10000/",winners_file), append = TRUE) 
  }
}

```




```{r}
total <- M*4
paste0("results_10000/",loop_winners_files)%>%
  sapply(read_csv, col_names = FALSE, simplify = FALSE) %>%
    bind_rows()%>%
  setNames("Team")%>%
  group_by(Team)%>%
  summarise(prob_win = n()/total)%>%
  arrange(desc(prob_win))%>%
  filter(prob_win > 0.001)%>%
  ggplot(aes(x=prob_win,y=reorder(Team,prob_win)))+
  geom_bar(stat = "identity") +
  geom_text(aes(label=round(prob_win,3)),position=position_dodge(width=0.9),hjust=-0.2)+
  labs(x="probability",
       y="",
       title = "Probability of winning the World Cup")+
  xlim(0,0.5)
```


# Completing submission template

```{r}

predict_knockout <- function(team1, team2){
  temp1 <- ifelse(team1 == "Republic of Ireland", "Ireland",
                  ifelse(team1 == "China PR", "China",
                         ifelse(team1 == "Korea Republic", "South Korea", team1)))
  temp2 <- ifelse(team2 == "Republic of Ireland", "Ireland",
                  ifelse(team2 == "China PR", "China",
                         ifelse(team2 == "Korea Republic", "South Korea", team2)))
  return(plogis(lambda[temp1] - lambda[temp2]))
}

output <- read_csv('submission-template.csv') %>%
  select(-starts_with('p')) %>%
  # Just use the market odds for the group stage
  left_join(estimated_df%>%
              mutate(TeamName1 = ifelse(TeamName1 == "Ireland", "Republic of Ireland",
                                        ifelse(TeamName1 == "China", "China PR",
                                               ifelse(TeamName1 == "South Korea", "Korea Republic", TeamName1))),
                     TeamName2 = ifelse(TeamName2 == "Ireland", "Republic of Ireland",
                                        ifelse(TeamName2 == "China", "China PR",
                                               ifelse(TeamName2 == "South Korea", "Korea Republic", TeamName2))))%>%
              select(group,
                     team1 = TeamName1,
                     team2 = TeamName2,
                     p_team1_win = p_ij,
                     p_team2_win = p_ji,
                     p_draw) %>%
              bind_rows(estimated_df %>%
                          mutate(TeamName1 = ifelse(TeamName1 == "Ireland", "Republic of Ireland",
                                                    ifelse(TeamName1 == "China", "China PR",
                                                           ifelse(TeamName1 == "South Korea", "Korea Republic", TeamName1))),
                                 TeamName2 = ifelse(TeamName2 == "Ireland", "Republic of Ireland",
                                                    ifelse(TeamName2 == "China", "China PR",
                                                           ifelse(TeamName2 == "South Korea", "Korea Republic", TeamName2))))%>%
                          select(group,
                                 team1 = TeamName2,
                                 team2 = TeamName1,
                                 p_team1_win = p_ji,
                                 p_team2_win = p_ij,
                                 p_draw))
  ) %>%
  # Predict knockout stage
  rowwise()%>%
  mutate(p_team1_win = ifelse(group == 'Knockout',
                              predict_knockout(team1,team2),
                              p_team1_win),
         p_team2_win = ifelse(group == 'Knockout',
                              predict_knockout(team2,team1),
                              p_team2_win),
         p_draw = ifelse(group == 'Knockout', 0, p_draw))%>%
  ungroup()
write.csv(output, 'output.csv')
output
```




